Cartograms with Rivers

### 2022-11-25 paper/section/evaluation.tex
draft: review

\subsection{User Study Tasks}

We formulate the following hypotheses to design our user study:

@@

We test these hypotheses using location based tasks. Given a standard choropleth on the left side of the screen and a cartogram on the right side, we ask user study participants to locate the given region in the choropleth in the cartogram. See \Cref{fig:task}.

@@

\bobgraph{Datasets:} We used the following datasets for our evaluation:

@@

\textbf{P3:} The participants are asked to complete 16 location tasks. The response and reaction time are recorded. These 16 CCGs were selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants were asked to complete a questionnaire that consists of Likert Scale questions and open-ended questions.

The materials for P1-P4 are available at \URL{https://osf.io/q39w7}.

@@

    {
        \begin{figure}[tb!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/accuracy.png}
            \caption{In the bar chart, the y-axis shows the average accuracy rate of locating CCGs with and without rivers, the x-axis shows the target CCGs.}
            \label{fig:task-acc}
        \end{figure}
    }

\bobgraph{Response Time:} \Cref{fig:task-rt} shows the average response time measured for four CCGs. The chart indicates that the introduction of rivers has slightly reduced the time needed to locate CCGs. The two groups show no statistically significant differences according to the two-sample unequal variance $t$-test performed, $t$(356.03) = 0.87, $p$ = 0.38.

    {
        \begin{figure}[tb!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/rt.png}
            \caption{In the bar chart, the y-axis shows the average response time for locating CCGs with and without rivers, the x-axis shows the target CCGs.}
            \label{fig:task-rt}
        \end{figure}
    }

@@

    {
        \begin{figure}[b!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/likert.png}
            \caption{The stacked bar chart shows the responses of Likert Scale questions.}
            \label{fig:likert}
        \end{figure}
    }

>>>>

\subsection{Study Hypothesis}

We formulate the following hypotheses to motivate our user study:

@@

We test these hypotheses using location based tasks. Given a standard choropleth on the left side of the screen and the equivalent cartogram on the right side, we ask user study participants to locate the given region in the choropleth (on the left) in the cartogram (on the right). See \Cref{fig:task}.

@@

\bobgraph{Datasets:} We used the following EHR datasets for our evaluation:

@@

\textbf{P3:} The participants are asked to complete 16 location tasks. The response and reaction time are recorded. These 16 CCGs are selected to avoid extreme cases (thus bias the result), in terms of size, color, and location.

\textbf{P4:} The participants are asked to complete a questionnaire that consists of Likert Scale questions and open-ended questions.

The materials for P1-P4 are available on the Open Science Foundation's website at \URL{https://osf.io/q39w7}.

@@

    {
        \begin{figure}[tb!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/accuracy.png}
            \caption{In the bar chart, the y-axis shows the average accuracy rate of locating CCGs with and without rivers for user study participants. The x-axis shows the target CCGs.}
            \label{fig:task-acc}
        \end{figure}
    }

\bobgraph{Response Time:} \Cref{fig:task-rt} shows the average time to located a CCG measured for four CCGs. The chart indicates that the introduction of rivers has slightly reduced the time needed to locate CCGs. The two groups show no statistically significant differences according to the two-sample unequal variance $t$-test performed, $t$(356.03) = 0.87, $p$ = 0.38. However, the average time to locate a given CCG is reduced for 3 out of the 4 CCGs users in the study.

    {
        \begin{figure}[tb!]
            \centering
            \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/rt.png}
            \caption{In the bar chart, the y-axis shows the average response time for locating CCGs with and without rivers for user study participants. The x-axis shows the target CCGs.}
            \label{fig:task-rt}
        \end{figure}
    }

@@

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth,keepaspectratio]{figure/evaluation/likert.png}
    \caption{The stacked bar chart shows the user study participant responses of Likert Scale questions.}
    \label{fig:likert}
\end{figure}
}

### 2022-11-25 paper/section/introduction.tex
draft: review

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \Cref{table:accuracy}). For this project, we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. Standard Demers cartograms are composed of uniform shape nodes. As such, this can reduce their legibility. We implement a new hybrid cartographic layout algorithm that combines rivers with the placement of the nodes representing geographical regions. Introducing rivers into the cartogram improves their legibility. To reduce geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real-world case study using Electronic Health Records (EHR) data to evaluate the result. We present a user study demonstrating its effectiveness.

>>>>

Among the four types of cartograms categorized by \citea{nusrat2016State} (contiguous, non-contiguous, rectangular, and Dorling), a trade-off between types of accuracy is made (See \Cref{table:accuracy}). For this project, we focus on non-contiguous cartograms like Demers cartograms, because they facilitate statistical comparison between regions and they can make good use of screen space, and comparison of regions is useful when studying EHR data. Building on Demers cartograms \cite{ian2002Cartogram}, we introduce and develop novel features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. Standard Demers cartograms are composed of uniform shape nodes. As such, this can reduce their legibility. We implement a new hybrid cartographic layout algorithm that combines rivers with the placement of the nodes representing geographical regions. Introducing rivers into the cartogram improves their legibility. To reduce geographical errors and make efficient use of screen space, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm to a real-world case study using Electronic Health Records (EHR) data to evaluate the result. We present a user study demonstrating its effectiveness.

### 2022-11-25 paper/section/main.tex
draft: review

\section{Cartograms with Rivers}
 {
  \begin{figure*}[tb!]
      \centering
      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{figure/flowchart.png}
      \caption{An overview of our hybrid layout algorithm. See \Cref{fig:flowchart-stalemate} for the logic of processing a stalemate. See also \Cref{alg:UpdateLayout} in \Cref{sec:{Cartograms with Rivers}} for more detail. For illustration purposes, we show the rendered views alongside the logical views representing the actual computation.}
      \label{fig:flowchart}
  \end{figure*}
 }

@@

 {
  \begin{figure}[tb!]
      \centering
      \includegraphics[width=\columnwidth,keepaspectratio]{figure/flowchart stalemate.png}
      \caption{A flowchart illustration of stalemate processing. See \Cref{subsec:{Detecting Stalemates}}, \Cref{fig:stalemate}, and \Cref{fig:corridor} for more detail.}
      \label{fig:flowchart-stalemate}
  \end{figure}
 }

@@

For all nodes $ \NxR $ that cross a river, $ \river $, we compute an average vector $ \vectorAvg $ used to translate $ \river $. Whenever a node, $ \node $, crosses a river , we store a vector $ \nodeVectorNNn $ that points in the direction of the translation. We then use a heuristic to translate $ \river $ using the average vector of node intersection $ \vectorAvg = \sum_{i=1}^j \frac{\Vector{\node_i\node_{i_t}}}{j} $, where $ j $ is the number of nodes intersecting the river. This step intends to create space for the next iteration of node translation without crossing a river. The procedure for translating riers is provided by \Cref{alg:TranslateRiver} in the Appendix.

When a river is translated by $ \vectorAvg $, this can trigger a scenario where the nodes are crossed by a translated river, denoted $ \RxN $. As a heuristic we also translate these nodes by $ \vectorAvg $. The reasoning behind this is that $ \vectorAvg $ indicates which direction the river needs to be translated in order to create space for nodes that are too crowded together. In practice, $ \vectorAvg $ is multiplied by a scaling factor $ \propto\vectorAvg $. Thus, we can influence how far a river, $ \river $, is translated in each iteration of the algorithm. We can use this to ensure smooth transitions between iterations.

@@

\bobgraph{Visibility:} The rendering visibility of various elements, including the choropleth shapefile, rivers, nodes, and node centroids, can be toggled on and off.

@@

\bobgraph{RIver Translation Behavior:} The behavior of rivers during the process can also be adjusted by the user: 1) Prevent river crossing: this option forbids nodes from crossing rivers; 2) Disable river translation: this option disables the translation of rivers. Both options are useful for generating different layouts and for observing the behavior of the hybrid layout algorithm.

\bobgraph{Corridor Length:} The user can define the length of a corridor, which is used to resolve stalemate situations. A longer corridor length allows more nodes to be translated when a stalemate occurs, but may result in a less accurate layout.

@@

{
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{figure/UI.png}
    \caption{A screenshot of the software. User options are provided to adjust the size, color mapping, visibility of nodes and rivers. Other options include the ability to control the overlap removal behavior: rivers can be static or dynamic.}
    \label{fig:overview}
\end{figure*}
}

>>>>

\section{Cartograms with Rivers}
 {
  \begin{figure*}[tb!]
      \centering
      \includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{figure/flowchart.png}
      \caption{An overview of our hybrid layout algorithm incorporating rivers. See also \Cref{alg:UpdateLayout} in \Cref{sec:{Cartograms with Rivers}} for more detail. See \Cref{fig:flowchart-stalemate} for the logic of processing a stalemate. For illustration purposes, we show the rendered views alongside the logical views representing the actual computation.}
      \label{fig:flowchart}
  \end{figure*}
 }

@@

 {
  \begin{figure}[tb!]
      \centering
      \includegraphics[width=\columnwidth,keepaspectratio]{figure/flowchart stalemate.png}
      \caption{A flowchart illustration of stalemate processing. See \Cref{subsec:{Process Stalemates}}, \Cref{fig:stalemate}, and \Cref{fig:corridor} for more detail.}
      \label{fig:flowchart-stalemate}
  \end{figure}
 }

@@

For all nodes $ \NxR $ that cross a river, $ \river $, we compute an average vector $ \vectorAvg $ used to translate $ \river $. Whenever a node, $ \node $, crosses a river , we store a vector $ \nodeVectorNNn $ that points in the direction of the translation. We then use a heuristic to translate $ \river $ using the average vector of node intersection $ \vectorAvg = \sum_{i=1}^j \frac{\Vector{\node_i\node_{i_t}}}{j} $, where $ j $ is the number of nodes intersecting the river. This step intends to create space for the next iteration of node translation without crossing a river. The procedure for translating rivers is provided by \Cref{alg:TranslateRiver}.

\begin{algorithm}[tbh!]
    \caption{Procedure to translate rivers.}\label{alg:TranslateRiver}
    \textbf{Input:} \\
    $ \nodeList \gets $ a list of $ \node $ representing CCGs \\

    \textbf{Local variables:} \\
    $ \riverList \gets $ a list of $ \river $ representing river features \\
    $ \node \gets $ a node is an object with the following properties: \\
    \-\hspace{1em} $ \node.x, \node.y, $ or $ \node(x,y) \gets $ the x and y coordinates of $ \node $ \\
    \-\hspace{1em} $ \nodePrevious \gets $ the previous position of $ \node $ \\
    \-\hspace{1em} $ \nodeFNOR \gets $ the translated position of $ \node $ \\
    $ j \gets $ the number of nodes intersecting $ \river $  \\

    \begin{algorithmic}[1]
        \Procedure{TranslateRiver}{$ \nodeList $}
        \ForEach {$ \river \in \riverList $}
        \State $ j \gets 0 $
        \State $ \riverVector \gets (0,0) $ \Comment{Hold the sum of vectors $ \nodeVectorCT $}
        \ForEach {$ \node \in \nodeList $}

        \If {$ \node(x,y) \neq \nodeFNOR(x,y) $}
        \If{ \Call{TestIntersection}{$ \nodeFNORLine $, $ \river $} $ = True $}
        \State $ \riverVector \gets \riverVector + \nodeVectorCT$
        \State $ j ++ $
        \EndIf

        \EndFor

        \State Translate river $ \river $ by the average vector $ \frac{\riverVector}{j} $
        \EndFor

When a river, $ \river $, is translated by $ \vectorAvg $, this can trigger a scenario where the nodes are crossed by a translated river, denoted $ \RxN $. As a heuristic we also translate these nodes by $ \vectorAvg $. The reasoning behind this is that $ \vectorAvg $ indicates which direction the river needs to be translated in order to create space for nodes that are too crowded together. In practice, $ \vectorAvg $ is multiplied by a scaling factor $ \propto\vectorAvg $. Thus, we can influence how far $ \river $ is translated in each iteration of the algorithm. We can use this to ensure smooth transitions between iterations.

@@

\bobgraph{Rendering Visibility:} The rendering visibility of various elements, including the choropleth shapefile, rivers, nodes, and node centroids, can be toggled on and off.

@@

\bobgraph{River Translation:} The behavior of rivers during the process can also be adjusted by the user: 1) Enable river crossing: this option forbids nodes from crossing rivers. Nodes cannot cross rivers by default; 2) Disable river translation: this option disables the translation of rivers, rivers are translated by default. Both options are useful for generating different layouts and for observing the behavior of the hybrid layout algorithm.

\bobgraph{Corridor Length:} The user can define the length of a corridor, which is used to resolve stalemate situations. A longer corridor length allows more nodes to be translated when a stalemate occurs, but may result in a less accurate layout. The default corridor length is three times the max node size.

@@

{
\begin{figure*}[tb!]
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{figure/UI.png}
    \caption{A screenshot of the software. User options are provided to adjust the size, color mapping, visibility of nodes and rivers. Other options include the ability to control the overlap removal behavior: rivers can be static or dynamic. See \Cref{subsec:{User Options}} for more on user options.}
    \label{fig:overview}
\end{figure*}
}

### 2022-11-25 paper/section/related.tex
draft: review

Our work extends the algorithm described by \citea{tong2018cartograms} which incorporates a static topological feature into Demers cartograms. Our work enhances that of \citeauthor{tong2018cartograms} in three ways. We introduce multiple features (rivers) into the layout (as opposed to a single), making topological features dynamic and further improving the legibility and geographical accuracy. We improve the algorithm for resolving stalemates.  The way we evaluate the cartograms is also different. \citeauthor{tong2018cartograms} count river crossings to evaluate error. Here our focus is on readability, thus we include a preliminary user study.

>>>>

Our work extends the algorithm described by \citea{tong2018cartograms} which incorporates a static topological feature into Demers cartograms. Our work enhances that of \citeauthor{tong2018cartograms} in multiple ways. First we introduce multiple features (rivers) into the layout, as opposed to a single river. Second, we make topological features dynamic and further improve the legibility and geographical accuracy. By the term \textit{dynamic}, we mean that the position of the rivers are updated as part of the layout algorithm. In previous work they are static and merely serve as boundaries. Third, we improve the algorithm for resolving stalemates. Finally, the way we evaluate the cartograms is also different. \citeauthor{tong2018cartograms} count river crossings to evaluate error. Here our focus is on readability, thus we include a preliminary user study.
